Improving disaggregation models of malaria incidence by ensembling non-linear models of prevalence

### 2019-12-12 tim-lucas-model-ensembles.tex
Minor author edits.
Tweaks.

\author[oxford]{Tim C. D. Lucas\corref{mycorrespondingauthor}}
\cortext[mycorrespondingauthor]{Corresponding author}
\ead{timcdlucas@gmail.com}
\author[oxford]{Anita K. Nandi}
\author[oxford]{Emma L. Collins}
\author[oxford]{Suzanne H. Keddie}
%\author[oxford]{Katherine A. Twohig}
\author[oxford]{Elisabeth G. Chestnutt}
\author[oxford]{Rosalind E. Howes}
%\author[oxford]{Michele Nguyen}
\author[oxford]{Susan F. Rumisha}
\author[oxford]{Andre Python}
\author[oxford]{Tasmin L. Symons}
\author[oxford]{Justin J. Millar}
\author[oxford]{Rohan Arambepola}
\author[oxford,idm]{Amelia Bertozzi-Villa}
\author[oxford]{Penelope Hancock}
\author[oxford]{Punam Amratia}
\author[oxford]{Katherine E. Battle}
\author[oxford]{Ewan Cameron}
\author[oxford,telethon,curtin]{Peter W. Gething}
\author[oxford]{Daniel J. Weiss}

@@

\begin{figure}[t!]
  \centering
  \includegraphics[width=0.8\textwidth]{figs/schematic2.pdf}
\caption{
  Schematic of the baseline disaggregation regression model and the two stage method. 
  Models are shown in yellow ovals, malaria data is shown in purple rectangles and covariates are shown in green rectangles.
  The baseline model uses aggregated incidence data and raw environmental covarates in a disaggregation regression model.
  In the two stage method, new covariates are created in stage 1 by training machine learning models on prevalence data.
  Predictions from these machine learning models are used as covariates in the stage 2 disaggregation regression.
  Only one of the two stage models (ML\textsubscript{l}) is shown for simplicity.
}
\label{f:schem}
\end{figure}
For each country specific dataset and for the global dataset we fitted 5 stage 1 models via \emph{caret} \citep{caret}: elastic net \citep{enet}, Random Forest \citep{wright2015ranger}, projection pursuit regression \citep{friedman1981projection}, neural networks \citep{nnet} and boosted regression trees (gradient boosted models, subsequently GBM) \citep{gbm}.
These models were fitted to both the full malaria prevalence dataset and to the regional subsets of the data.
Our response variable was prevalence and we weighted the data by sample size (i.e.\thinspace the number of people tested for malaria in each survey).
We used the raw environmental covariates described above as covariates in these machine learning models.
This process can therefore be seen as creating non-linear transformations of the raw covariates that are hopefully better correlated with malaria incidence than the raw environmental covariates are.
For each model we ran five-fold cross-validation to select hyperparameters using random search for Random Forest and boosted regression trees and grid search for the other models. 
Root mean square error (RMSE) was used to select the best performing model.
Predictions from these models were then made across each country respectively.
These predictions were empirical logit transformed so that they were on the linear predictor scale of the disaggregation regression model.
An empirical logit was used rather than a standard logit as there were many predictions of exactly zero.
These predicted surfaces were subsequently used as covariates in the stage 2 dissaggregation regression models (Figure~\ref{f:schem}).
The experiments that follow in this paper assess the improved performance of disaggregation regression models when using the predictions from these machine learning models as covariates as compared to a baseline disaggregation regression model using only the raw environmental covariates.

@@

@@

All experiments involve comparing predictive performance of stage two disaggregation regression models when different combinations of covariates are used.
We used the raw environmental covariates, centered and standardised, as a baseline stage two model (Figure~\ref{f:schem}).
We subsequently refer to this model as Enviro.
We then performed two experiments.
In the first (\emph{i}) we tested whether using predictions from stage one machine learning models trained on local (i.e. within-country) prevalence data improved predictions (Figure~\ref{f:schem}).
In the second (\emph{ii}) we tested whether using predictions from stage one machine learning models trained on global prevalence data improved predictions.
These combinations of covariates are summarised in Table~\ref{t:models}.
Our primary performance metric was correlation between observed and predicted aggregated  data.
We also examined the calibration of models by calculating the proportion of held out data that were within their 80\% credible intervals.

In experiment \emph{i} we compared the baseline to two models that used locally trained machine learning models.
Firstly, we combined the predictions from the machine learning models and the environmental coviarates in one model (subsequantly called Enviro + ML\textsubscript{l}).
In this model therefore, the environmental covariates are effectively used twice, once in their raw form and once in transformed form (i.e. the predictions from the machine learning models).
Secondly, we used only the predictions from the machine learning models (Figure~\ref{f:schem}) trained on local data (subsequantly called ML\textsubscript{l}).
As the environmental covariates are used as covariates in the machine learning models, this model is still ultimately driven by the values of the raw environmental covariates.
However, in this model we are only using the environmental covariates in the transformed space learned by the machine learning models.

In experiment \emph{ii} we compared the baseline model to two models that use predictions from machine learning models trained on the global dataset of prevalence surveys.
In the first, we used only predictions from the machine learning models trained on the global data (ML\textsubscript{g}).
In the second, we combined predictions from the machine learning models trained on regional data and predictions from the machine learning models trained on global data (ML\textsubscript{l} + ML\textsubscript{g}).

>>>>

\author[oxford]{Tim C. D. Lucas\corref{mycorrespondingauthor}}
\cortext[mycorrespondingauthor]{Corresponding author}
\ead{timcdlucas@gmail.com}
\author[oxford]{Anita K. Nandi}
\author[oxford]{Suzanne H. Keddie}
%\author[oxford]{Katherine A. Twohig}
\author[oxford]{Elisabeth G. Chestnutt}
\author[oxford]{Rosalind E. Howes}
%\author[oxford]{Michele Nguyen}
\author[oxford]{Susan F. Rumisha}
\author[oxford]{Rohan Arambepola}
\author[oxford,idm]{Amelia Bertozzi-Villa}
\author[oxford]{Andre Python}
\author[oxford]{Tasmin L. Symons}
\author[oxford]{Justin J. Millar}
\author[oxford]{Punam Amratia}
\author[oxford]{Penelope Hancock}
\author[oxford]{Katherine E. Battle}
\author[oxford]{Ewan Cameron}
\author[oxford,telethon,curtin]{Peter W. Gething}
\author[oxford]{Daniel J. Weiss}

@@

\begin{figure}[t!]
  \centering
  \includegraphics[width=0.8\textwidth]{figs/schematic2.pdf}
\caption{
  Schematic of the baseline disaggregation regression model and the two stage method. 
  Models are shown in yellow ovals, malaria data is shown in purple rectangles and covariates are shown in green rectangles.
  The baseline model uses aggregated incidence data and raw environmental covarates in a disaggregation regression model.
  In the two stage method, new covariates are created in stage 1 by training machine learning models on prevalence data.
  Predictions from these machine learning models are used as covariates in the stage 2 disaggregation regression.
  Only one of the two stage models (ML\textsubscript{l}) is shown for simplicity.
}
\label{f:schem}
\end{figure}

For each country specific dataset and for the global dataset we fitted 5 stage 1 models via \emph{caret} \citep{caret}: elastic net \citep{enet}, Random Forest \citep{wright2015ranger}, projection pursuit regression \citep{friedman1981projection}, neural networks \citep{nnet} and boosted regression trees (gradient boosted models, subsequently GBM) \citep{gbm}.
These models were fitted to both the full malaria prevalence dataset and to the regional subsets of the data.
Our response variable was prevalence and we weighted the data by sample size (i.e.\thinspace the number of people tested for malaria in each survey).
We used the raw environmental covariates described above as covariates in these machine learning models.
This process can therefore be seen as creating non-linear transformations of the raw covariates that are hopefully better correlated with malaria incidence than the raw environmental covariates are.
For each model we ran five-fold cross-validation to select hyperparameters using random search for Random Forest and boosted regression trees and grid search for the other models. 
Root mean square error (RMSE) was used to select the best performing model.
Predictions from these models were then made across each country respectively.
These predictions were empirical logit transformed so that they were on the linear predictor scale of the disaggregation regression model.
An empirical logit was used rather than a standard logit as there were many predictions of exactly zero.
These predicted surfaces were subsequently used as covariates in the stage 2 dissaggregation regression models (Figure~\ref{f:schem}).
The experiments that follow in this paper assess the improved performance of disaggregation regression models when using the predictions from these machine learning models as covariates as compared to a baseline disaggregation regression model using only the raw environmental covariates.

@@

All experiments involve comparing predictive performance of stage two disaggregation regression models when different combinations of covariates are used.
We used the raw environmental covariates, centered and standardised, as a baseline stage two model (Figure~\ref{f:schem}).
We subsequently refer to this model as Enviro.
We then performed two experiments.
In the first (\emph{i}) we tested whether using predictions from stage one machine learning models trained on local (i.e. within-country) prevalence data improved predictions (Figure~\ref{f:schem}).
In the second (\emph{ii}) we tested whether using predictions from stage one machine learning models trained on global prevalence data improved predictions.
Our primary performance metric was correlation between observed and predicted aggregated  data.
We also examined the calibration of models by calculating the proportion of held out data that were within their 80\% credible intervals.

In experiment \emph{i} we compared the baseline to two models that used locally trained machine learning models (Table~\ref{t:models}).
Firstly, we combined the predictions from the machine learning models and the environmental coviarates in one model (subsequently called Enviro + ML\textsubscript{l}).
In this model therefore, the environmental covariates are effectively used twice, once in their raw form and once in transformed form (i.e. the predictions from the machine learning models).
Secondly, we used only the predictions from the machine learning models (Figure~\ref{f:schem}) trained on local data (subsequantly called ML\textsubscript{l}).
As the environmental covariates are used as covariates in the machine learning models, this model is still ultimately driven by the values of the raw environmental covariates.
However, in this model we are only using the environmental covariates in the transformed space learned by the machine learning models.

@@

In experiment \emph{ii} we compared the baseline model to two models that use predictions from machine learning models trained on the global dataset of prevalence surveys (Table~\ref{t:models}).
In the first, we used only predictions from the machine learning models trained on the global data (ML\textsubscript{g}).
In the second, we combined predictions from the machine learning models trained on regional data and predictions from the machine learning models trained on global data (ML\textsubscript{l} + ML\textsubscript{g}).
